A properly labelled proof system for intuitionistic modal logics

### 2019-09-07 caleidoscope/caleidoscope.tex
starting the intro

\documentclass[twoside]{aiml18}

\usepackage{aiml18macro}

\usepackage[noxy]{virginialake}                        
\usepackage{lmodern}
\usepackage{amssymb,amsmath} 
\usepackage{mathtools}
\usepackage{enumerate}
\usepackage{graphicx}
\usepackage{bm}
\usepackage{xcolor}
\usepackage{pgf}
\usepackage[all]{xy}
\usepackage{tikz}
%\usepackage{hyperref}
\usepackage{pgffor}

@@

%%% General
\input{macros}
\newcommand{\B}{\mathcal{B}}
\newcommand*{\A}{\mathcal{A}}

%%% Systems
\newcommand*{\ax}[1]{\mathsf{#1}}
\newcommand*{\kax}[1][]		{\ax{k_{#1}}}
\newcommand*{\lab}{\mathsf{lab}}
\newcommand*{\IK}{\mathsf{IK}}
\newcommand*{\labIKp}{\lab\IK_{\le}}

%%% Connectives
\newcommand*{\NOT}{\neg}
\newcommand*{\AND}{\mathbin{\wedge}}
\newcommand*{\TOP}{\mathord{\top}}
\newcommand*{\OR}{\mathbin{\vee}}
\newcommand*{\BOT}{\mathord{\bot}}
\newcommand*{\IMP}{\mathbin{\scalebox{.9}{\raise.2ex\hbox{$\supset$}}}}

@@

%%% Labelled sequents
\newcommand*{\labels}[2]{{\color{blue}{#1}\:\colon}{#2}}
\newcommand{\SEQ}{\Rightarrow}
\newcommand*{\DD}{\mathcal{D}}
\newcommand*{\rn}[1]  {\ensuremath{\mathsf{#1}}}
\newcommand*{\invr}[1]{#1^\bullet}
\newcommand*{\rel}{R}

%%% Labelled rules
\newcommand*{\labrn}[2][]  {\rn{#2}_{#1}}%^{\lab}}}
\newcommand*{\rlabrn}[2][]  {\rn{#2}_\rn{R#1}}%^\lab}}
\newcommand*{\llabrn}[2][]  {\rn{#2}_\rn{L#1}}%^\lab}}
%
\newcommand*{\brsym}{\boxtimes}%\mathord{\scalebox{.8}{$\blacksquare$}}}
\newcommand*{\diasym}{\diamondplus}%\mathord{\blacklozenge}}
%
\newcommand*{\boxbrn}[1]{\brsym_\rn{#1}}%^{\lab}}}
\newcommand*{\diabrn}[1]{\diasym_\rn{#1}}

@@

@@

\begin{frontmatter}
  \title{Decomposing labelled proof theory for intuitionistic modal logic}
  \author{Sonia Marin}

  \todo{New university?}
  \address{IT-Universitetet i K{\o}benhavn \\ Denmark }
 \author{Marianela Morales}

% \marianela{I think that now I have the same as Lutz, right?} No, the other order...
 \address{ LIX, \'Ecole Polytechnique  \& Inria Saclay\\ France}
   \author{Lutz Stra{\ss}burger}
 \address{Inria Saclay \& LIX, \'Ecole Polytechnique \\ France}

@@

   We present a labelled sequent system for intuitionistic modal logic equipped 
   with two relation symbols, one for the accessibility
   relation associated with the Kripke semantics for modal logics and
   one for the preorder relation associated with the Kripke semantics
   for intuitionistic logic. 
   %
   Thus, it is in close correspondence with birelational Kripke semantics for intuitionistic modal logic and can encompass a wider range of intuitionistic logics than state-of-the-art labelled systems.

@@

  \begin{keyword}
  Intuitionistic modal logic, Labelled sequents, Proof theory.
  \end{keyword}
 \end{frontmatter}

\section{Introduction}
\todo{We should work on the introduction}

@@

>>>>

\documentclass[a4paper]{article}

%\usepackage{aiml18macro}

\usepackage[margin=3cm]{geometry}
\usepackage[noxy]{virginialake}                        
\usepackage{lmodern}
\usepackage{amssymb,amsmath} 
\usepackage{mathtools}
\usepackage{enumerate}
\usepackage{graphicx}
\usepackage{bm}
\usepackage{xcolor}
\usepackage{pgf}
\usepackage[all]{xy}
\usepackage{tikz}
%\usepackage{hyperref}
\usepackage{pgffor}

@@

%%% General
\input{macros}
%\newcommand{\B}{\Digamma}
\newcommand{\B}{\mathcal{R}}
%\newcommand{\B}{\Xi}
%\newcommand{\B}{\mathcal{B}}
\newcommand*{\A}{\mathcal{A}}

%%% Systems
\newcommand*{\ax}[1]{\mathsf{#1}}
\newcommand*{\kax}[1][]		{\ax{k_{#1}}}
\newcommand*{\lab}{\mathsf{lab}}
\newcommand*{\IK}{\mathsf{IK}}
\newcommand*{\K}{\mathsf{K}}
\newcommand*{\labIKp}{\lab\IK_{\le}}
\newcommand*{\ISfour}{\mathsf{IS4}}
\newcommand*{\Sfour}{\mathsf{S4}}

\newcommand{\fourax}{\mathsf{4}}
%%% Connectives
\newcommand*{\NOT}{\neg}
\newcommand*{\AND}{\mathbin{\wedge}}
\newcommand*{\TOP}{\mathord{\top}}
\newcommand*{\OR}{\mathbin{\vee}}
\newcommand*{\BOT}{\mathord{\bot}}
\newcommand*{\IMP}{\mathbin{\scalebox{.9}{\raise.2ex\hbox{$\supset$}}}}

@@

%%% Labelled sequents
\newcommand*{\fm}[1]{{\color{notgreen}{#1}}}
\newcommand*{\lb}[1]{{\color{blue}{#1}}}
\newcommand*{\labels}[2]{\lb{#1}\mathord{:}\fm{#2}}
\newcommand*{\accs}[2]{\lb{#1}R\lb{#2}}
\newcommand*{\futs}[2]{\lb{#1}\le{\color{blue}{#2}}}
\newcommand{\SEQ}{\Longrightarrow}
\newcommand*{\DD}{\mathcal{D}}
\newcommand*{\rn}[1]  {\ensuremath{\mathsf{#1}}}
\newcommand*{\invr}[1]{#1^\bullet}
\newcommand*{\rel}{R}

\newcommand{\proviso}[1]{\mbox{\scriptsize #1}}
%%% Labelled rules
\newcommand*{\labrn}[2][]  {\rn{#2}_{#1}}%^{\lab}}}
\newcommand*{\rlabrn}[2][]  {\rn{#2}_\rn{R#1}}%^\lab}}
\newcommand*{\llabrn}[2][]  {\rn{#2}_\rn{L#1}}%^\lab}}
%
\newcommand*{\brsym}{\boxtimes}%\mathord{\scalebox{.8}{$\blacksquare$}}}
\newcommand*{\diasym}{\diamondplus}%\mathord{\blacklozenge}}
%
\newcommand*{\boxbrn}[1]{\brsym_\rn{#1}}%^{\lab}}}
\newcommand*{\diabrn}[1]{\diasym_\rn{#1}}

@@

%%%%%%%%%%%%%%%%%

\newcommand{\qquand}{\qquad\mbox{and}\qquad}
\newcommand{\quadcm}{\rlap{\quad,}}

@@

%\begin{frontmatter}
  \title{A properly labelled proof system for intuitionistic modal logics}
  \author{Sonia Marin \and %}
%  \todo{New university?}
%  \address{IT-Universitetet i K{\o}benhavn \\ Denmark }
%    \author{
      Marianela Morales %}
 \and
% \marianela{I think that now I have the same as Lutz, right?} No, the other order...
% \address{ LIX, \'Ecole Polytechnique  \& Inria Saclay\\ France}
%      \author{
        Lutz Stra{\ss}burger }
% \address{Inria Saclay \& LIX, \'Ecole Polytechnique \\ France}

  \maketitle

@@

   In this paper we present a labelled sequent system for
   intuitionistic modal logics such that there is not only one, but
   two relation symbols for the labels: one for the accessibility
   relation associated with the Kripke semantics for modal logics and
   one for the preorder relation associated with the Kripke semantics
   for intuitionistic logic.
   %%
   This puts our system in close correspondence with the standard
   birelational Kripke semantics for intuitionistic modal logic. As a
   consequence it can encompass a wider range of intuitionistic modal
   logics than standard labelled systems.
   %%
   We also show an internal cut elimination proof for our system.

@@

  \begin{keyword}
  Intuitionistic modal logic, labelled sequents, cut elimination, proof theory.
  \end{keyword}
% \end{frontmatter}

  %\maketitle
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\todo{get bib-file right}

\section{Introduction}\label{sec:intro}

Since their introduction in the 1980s by Gabbay~\cite{Gabbay},
\emph{labelled proof calculi} have been widely used by proof theorist
to give sound, complete and cut-free deductive system to a broad range
of logics. Unlike so-called \emph{internal calculi}, like
hyper-sequents~\cite{avron} or nested
sequents~\cite{kashima,brunnler,pog} or 2-sequents~\cite{masini},
labelled calculi have the advantage of beeing more uniform and
accommodating more logics.

Standard labelled sequent calculi attach to every formula $A$ a label
$x$, witten as $\labels xA$, and additionally admit so called
\emph{relational atoms} of the form $\accs xy$ where $\rel$ is a
binary relation symbol. These calculi work best for logics with
standard Kripke semantics, as in this case the relation $\rel$ is used
to encode the accessibility relation in the Kripke models, and the
frame condition corresponding to the desired logic can be directly
encoded as inference rules. Prominent examples are classical modal
logics and intuitionistic propositional logic, where, e.g., the frame
condition of transitivity ($\forall \lb {xyz}.\:\accs xy\AND \accs yz\IMP\accs xz$), can be straightforwardly translated into the
inference rule
\begin{equation}
  \label{eq:Rtrans}
  \vlinf{\rn{trans}}{}{\B, \accs xy, \accs yz, \Left \SEQ \Right}{
    \B, \accs xy, \accs yz, \accs xz, \Left \SEQ \Right}
\end{equation}
where $\B$ stands for a set of relation atoms, and $\Gamma$ and
$\Delta$ for sets \lutz{or multisets?} of labelled formulas.

However,
in this paper we are concerned with \emph{intuitionistic modal
  logics}, whose Kripke semantics is based on \emph{birelational}
frames, i.e., they have two binary relations instead of one: one
relation $R$ that corresponds to the accessibility relation in
ordinary Kripke frames for modal logics, and a relation $\le$ that
corresponds to the future relation in ordinary Kripke frames for
intuitionistic logic. Consequently, standard labelled systems for
these logics have certain shortcomings:
\begin{enumerate}
\item The transitivity rule in~\eqref{eq:Rtrans} corresponds to both
  versions of the $\fourax$-axiom
  \begin{equation}
    \label{eq:4ax}
    \fm{\BOX A\IMP \BOX\BOX A}
    \qquand
    \fm{\DIA\DIA A\IMP\DIA A}
    \quadcm
  \end{equation}
  as they are equivalent in classical
  modal logic. However, in intuiotionistic modal logic they are not
  equivalent, and even though the logic $\ISfour$, i.e., the
  intuitionistic version of the modal logic $\Sfour$, contains both
  axioms, they can be added independently to the logic $\IK$ (the
  intuitionistic version of $\K$). However, the proof theory of these
  new logics has not been studied before, as no existing labelled (or
  label-free) proof system can handle them, even though the
  corresponding frame conditions
  \begin{equation}
    \label{eq:4frame}
    \forall\lb{xyz}.\:\accs xy\AND \accs yz\IMP(\exists\lb{x'}.\:\futs x{x'}\AND\accs {x'}z)
    \qquand
    \forall\lb{xyz}.\:\accs xy\AND \accs yz\IMP(\exists\lb{z'}.\:\futs z{z'}\AND\accs {x}{z'})
  \end{equation}
  have already been studied
  in~\cite{Plotkin}.\lutz{check: did they get that wrong in  \cite{Plotkin}?}
\item The correspondence between the syntax and the semantics is not
  as clean as one would expect. As only the $R$-relation (and not the
  $\le$-relation) of the frame is visible in an ordinary labelled
  sequent, we only have that a sequent $\Gamma$ is provable if and
  only if is is valid in all \emph{graph-consistent}\footnote{This
    means that every layer in the model can be lifted to any future of
    any world in that layer.}\lutz{check terminology} models, as
  already observed in by Simpson in his PhD thesis~\cite{simpson:phd}
  and considered as a very unsatisfactory situation\lutz{check what he
    said exactly} (see also \cite{mar:str:indexed}).
\end{enumerate}
In order to address these two issues we propose here to extend the
notion of labelled sequent by also allowing relational atoms of the
form $\futs xy$. Consequently, we can easily translate the frame
conditions in~\eqref{eq:4frame} into inference rules:
\begin{equation}
  \label{eq:4rules}
  \vlinf{\rn{4^\BOX}}{\proviso{$\lb{x'}$ fresh}}{\B, \accs xy, \accs yz, \Left \SEQ \Right}{
    \B, \accs xy, \accs yz, \accs {x'}z, \futs x{x'},\Left \SEQ \Right}
  \qquand
  \vlinf{\rn{4^\DIA}}{\proviso{$\lb{z'}$ fresh}}{\B, \accs xy, \accs yz, \Left \SEQ \Right}{
    \B, \accs xy, \accs yz, \accs {x}{z'}, \futs z{z'},\Left \SEQ \Right}  
\end{equation}

\section{Introduction(old version, to be deleted)}
\todo{We should work on the introduction}

@@

\sonia{
  \begin{itemize}
  \item internal vs externl calculi
  \item extesion to axioms are truly intuitionistic (not just the
    classical versions as in Simpson style)
  \item connection with semantics compared to graph homomorphism
    restiction in Simpson and ``indexed nested systems''
  \end{itemize}
}
